Constrained, yet diverse: Phylogenetic and morphological filtering in epiphytic ferns

### 2019-10-18 ms/manuscript.Rmd
Update MS
Restructure discussion

Here, we show for the first time using a broadly sampled phylogenetic dataset that multiple morphological traits related to epiphytic growth are evolutionarily conserved in ferns. We find that epiphytic ferns are constrained by size, having consistently smaller fronds and rhizomes than terrerstial species. However, contrary to our hypothesis, epiphytic fern communities appear to be more diverse than terrestrial communities. This suggests that environmental filtering plays a strong role in limiting plant size, while allowing for diversification across other trait axes in epiphytic ferns.

### Effect of Phylogeny on Analysis of Epiphytic Traits

Although several other studies have addressed the relationship of functional traits with growth habit in ferns [e.g., @Hietz1998; @Kluge2007; @WatkinsJr.2010], few have accounted for evolutionary history. We find that growth habit shows phylogenetic signal, as do several of the traits associated with it. In one of the few previous studies to analyze phylogenetic signal in functional traits of ferns, @Zhang2014b found that only one of 16 traits showed significant phylogenetic signal. The only trait in common between the current study and that of @Zhang2014b is SLA, which did not show phylogenetic signal in either study. Given the high degree of phylogenetic signal we observe in most traits and that at least some of the traits between our two studies are likely to be correlated, it is surprising that @Zhang2014b did not detect phylogenetic signal in more traits. However, @Zhang2014b only used \K, which tended to indicate less phylogenetic signal than $\lambda$ in our study. In simulation studies, $\lambda$ has lower error (both Type I and Type II) than \K, which can be inconsistent unless the strength of BM is great [@Munkemuller2012]. Our results indicate the importance of including a phylogenetic component, and including multiple measures of phylogenetic signal, when analyzing traits related to epiphytic growth in ferns.

### Sporophyte Traits

Epiphytes were markedly smaller than terrestrial species as measured by stipe and frond length, frond width, and rhizome diameter at both the species and community level (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)). Our results at the local scale are corroborated by @Testo2018a, who also observed abrupt size reduction associated with epiphytic growth in a globally sampled fern phylogeny. Stipe length in particular may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance [@WatkinsJr.2010]. Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns [@WatkinsJr.2010; @Creese2011]. However, given that these were all correlated with each other in our study, and similar traits have been shown to scale together previously [@Arcand2008; @Creese2011], careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

```{r sla-numbers, results = "hide"}
# First smallest SLA: Humata_anderssonii
smallest_sla_1 <- fern_traits %>% arrange(sla) %>% slice(1) %>% pull(sla) %>% signif(3)
assert_that(fern_traits %>% arrange(sla) %>% slice(1) %>% pull(species) == "Humata_anderssonii")

# Second smallest SLA: Ctenopterella_blechnoides
smallest_sla_2 <- fern_traits %>% arrange(sla) %>% slice(2) %>% pull(sla) %>% signif(3)
assert_that(fern_traits %>% arrange(sla) %>% slice(2) %>% pull(species) == "Ctenopterella_blechnoides")
```

In contrast to their clearly smaller size, epiphytes were remarkably diverse in other traits. Epiphytes tended to have less dissected fronds with fewer pinna than terrestrial species (\tab \@ref(tab:pics)), but the range of values for these traits was quite wide at the community level (error bars in \fig \@ref(fig:cwm-div)). Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)). Specific leaf area is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a]. We expected that epiphytes would have lower SLA, given their more extreme growing conditions. However, the parallel relationship in CWM values of SLA with elevation in both epiphytes and terrestrial species (\fig \@ref(fig:cwm-div)) indicates that SLA may be responding more strongly to temperature in both epiphytes and terrestrial species, while only being weakly modified by humidity.

### Gametophyte Traits

Only gametophyte morphotype was found to have a significant association with epiphytic growth. Morphotype was previously observed to correlate with life history and habitat in ferns: terrestrial species tend to have short-lived ($< \SI{1}{\year}$), cordate gametophytes that establish following disturbance and rapidly produce sporophytes, whereas epiphytic species tend to have non-cordate (\eg, elongate or filamentous) gametophytes that persist over multiple growing seasons and produce sporophytes more slowly [@WatkinsJr.2007a; @Farrar2008]. The complex, three-dimensional structure of non-cordate gametophytes slows their drying rates, and may be an adaptation to drought-prone epiphytic habitats [@WatkinsJr.2007c; @Pittermann2013], and promote out-crossing [@Farrar2008]. Our study is the first to our knowledge to demonstrate that the correlation between non-cordate morphology and epiphytism in ferns is significant while controlling for phylogeny. @Sundue2015 found that non-cordate morphology is significantly correlated with increased diversification rates within Polypodiaceae, consistent with the hypothesis that this trait promoted diversification within the epiphytic niche.

We did not find that gemmae production was linked with epiphytism. Gemmae are asexual propagules that allow gametophytes to persist over multiple growing seasons and attain large population sizes [@Farrar2008]. Gemmae production should be a useful trait in an epiphytic context because it may allow gametophytes to increase their surface area and hedge against population loss due to extreme drying events [@Farrar2008]. Gemmae can also function sexually by producing antherida in the presence of antheridiogen secreted by other gametophytes [@Emigh1977], which may confer a selective advantage in the epiphytic environment by enabling outcrossing [@Dassler2001]. It is possible that some, but not all, epiphytic ferns rely on such strategies, and that they are not under similarly strong selection as other traits such as leaf size and gametophyte morphotype. Interestingly, when we classified 'widespread' species as those with gametophytes occurring above or below the minimum or maximum elevational range of conspecific sporophytes using the data of @Nitta2017, we found that gemmae production is correlated with widespread growth (log likelihood dependent model = `r correlated_evo_test_gemmae$dependent.logL %>% round(1) %>% as.numeric`, log likelihood independent model = `r correlated_evo_test_gemmae$independent.logL %>% round(1) %>% as.numeric`, \pval = `r correlated_evo_test_gemmae$P %>% round(3) %>% as.numeric`, Pagel's test of correlated evolution). Thus, it seems that gemmae may function in long-term gametophyte persistence, but are not required for epiphytic growth.

Presence of hairs and glands has strong phylogenetic signal (\tab \@ref(tab:phylosig-binary)), and these traits are correlated with certain taxonomic groups [@Stokey1951; @Nayar1971]. Although hairs have been hypothesized to reduce rates of drying and increase water uptake after desiccation [@Kappen2007], we did not detect a correlation between hair production and epiphytic growth. Similarly, production of glands does not appear to be correlated with epiphytic growth. Wax-secreting glands are prevalent in both sporophytes and gametophytes of many notholaenid ferns, a xeric-adapted clade, but their functional significance has not been established [@Johnson2012]. Further tests are needed to determine if such glands serve a functional role in preventing water loss in desert ferns and epiphytes.

### Environmental Filtering

@@

We obtained partial support for our hypothesis that epiphytic fern communities would be more phylogenetically clustered than terrestrial ones; this was observed for MNTD, but not MPD (\fig \@ref(fig:comm-div)). This makes sense given that most epiphytes have radiated within five major clades [@Schuettpelz2009]; each epiphytic community we sampled tends to include closely-related species from two or more epiphytic clades, but has a similar overall level of phylogentic diversity as terrestrial communities. However, we did not detect significant trends in phylogenetic community structure of epiphytic and terrestrial communities along the elevational gradient. It is possible that the scale of the current study (from \SIrange{200}{1200}{\meter}) prevented the detection of such trends. When the plots in this study were analyzed together with additional plots from Tahiti reaching a maximum altitude of just over \SI{2000}{\meter}, @Nitta2017 detected increasing phylogenetic clustering in fern sporophyte communities (epiphytes and terrestrial species together) with elevation. @Kluge2011 also compared phylogenetic and trait diversity between epiphytic and terrestrial ferns over an elevational gradient from \SIrange{100}{3400}{\meter} in Costa Rica, and found trait overdispersion at mid-elevations for epiphytes but no trends in phylogenetic diversity for either group. They accounted this to evolutionary plasticity in traits, but did not quantify the degree of phylogenetic signal in the traits they studied.

Results for functional community diversity contradicted our hypotheses; epiphytic communities were more functionally diverse than terrestrial as measured by both MPD and MNTD (\fig \@ref(fig:comm-div)). However, this result is dependent on the traits used to measure functional diversity; when the same analyses were done using only size-related traits (scaled, untransformed stipe length, frond length and width, and rhizome diameter), the opposite result was observed (\fig S1). Futhermore, terrestrial species show a much wider range of values for size traits (\fig \@ref(fig:cwm-div)), and neither functional group appears to be restricted to a smaller area in trait space (\fig \@ref(fig:pca)). This suggests that epiphytic ferns are subject to environmental filtering for size-related traits, while simultaneously diversifying along other trait axes. This could be maintained by niche partitioning within epiphytic habitats, which may be more diverse than terrestrial for a given area [@Rojas-Hernandez2018]. Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk [@Hietz1995]. The vertical stratification of epiphytic species has been well documented in ferns [\eg, @Parra2009; @WatkinsJr.2006a; @Acebey2017] and broader studies [\eg, @Kromer2007; @Zotz2007]. The goal of our study was to quantify differences between epiphytic and terrestrial fern communities; future research should focus on bringing comparative phylogenetic and community ecological methods to bear on the large body of work characterizing differences within epiphytic microhabitats.

### Conclusion and Future Directions

We have demonstrated that growth habit, and several traits associated with it, show significant phylogenetic signal in ferns. Furthermore, we find that epiphytes are constrained by size while being remarkably diverse in other traits. Finally, we observed drier conditions in the canopy relative to terrestrial habitats. Taken together, our results support a scenario of environmental filtering acting to limit size and phylogenetic diversity in epiphytic ferns, while simulataneously enabling diversification along other trait axes. The generality of this pattern could be further tested by investigating the functional traits of other groups with large epiphytic radiations such as orchids and bromeliads.

Recent studies [e.g., @Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a] have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by @Schuettpelz2009 by failing to detect increased rates of diversification associated with epiphytic growth. Epiphytism itself is likely a complex trait requiring the evolution of multiple associated traits, and the timing and order of the evolution of such traits may be related to the complex diversification dynamics observed in these studies. Small size seems to be a prerequisite for epiphytic growth, as indicated both by our results and the global study of @Testo2018a. Here, we identify several traits in addition to plant size (\eg, pinna number, frond dissection, gametophyte morphotype) that may also belong to the suite of epiphytic traits. We hope future studies will build on our results by analyzing traits involved in epiphytic growth other than body size across densely sampled phylogenies to dissect how each may relate to diversification and growth habit. Additionally, future studies should focus on determining in more detail how niche space is partitioned within epiphytic vs. terrestrial habitats, which is clearly a major dividing line for ferns.

>>>>

Here, we provide the first study to our knowledge to combine comparative methods with phylogenetic community ecology to analyze the evolutionary implications of epiphytic growth in ferns.
We find that epiphytic ferns are constrained in size, having consistently smaller fronds and rhizomes than terrestrial species.
In contrast, epiphytic ferns vary greatly in other traits, leading epiphytic fern communities to be more functionally diverse overall than terrestrial communities.
Taken together, our results suggest that environmental filtering plays a strong role in limiting plant size, while simultaneously allowing for diversification across other trait axes in epiphytic ferns.

### Convergence in small size and non-cordate morphotype

Epiphytes were markedly smaller than terrestrial species as measured by stipe and frond length, frond width, and rhizome diameter at both the species and community level (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)).
Our results at the local scale are corroborated by @Testo2018a, who also observed abrupt size reduction associated with epiphytic growth in a globally sampled fern phylogeny.
Stipe length in particular may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance [@WatkinsJr.2010].
Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns [@WatkinsJr.2010; @Creese2011].
However, given that stipe and frond length, frond width, and rhizome diameter were correlated, and similar traits have been shown to scale together previously [@Arcand2008; @Creese2011], careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

Another clear pattern to emerge from our study is the association between epiphytism and non-cordate (\eg, elongate or filamentous) morphotype.
Morphotype was previously observed to correlate with life history and habitat in ferns: terrestrial species tend to have short-lived ($< \SI{1}{\year}$), cordate gametophytes that establish following disturbance and rapidly produce sporophytes, whereas epiphytic species tend to have non-cordate gametophytes that persist over multiple growing seasons and produce sporophytes more slowly [@WatkinsJr.2007a; @Farrar2008].
The complex, three-dimensional structure of non-cordate gametophytes slows their drying rates, and may be an adaptation to drought-prone epiphytic habitats [@WatkinsJr.2007c; @Pittermann2013], and promote out-crossing [@Farrar2008].
Our study is the first to our knowledge to demonstrate that the correlation between non-cordate morphology and epiphytism in ferns is significant while controlling for phylogeny.
@Sundue2015 found that non-cordate morphology is significantly correlated with increased diversification rates within Polypodiaceae, consistent with the hypothesis that this trait promoted diversification within the epiphytic niche.

### Remarkable diversity in other traits

In contrast to their clearly smaller size and non-cordate morphotypes, epiphytes were remarkably diverse in other traits.
Epiphytes tended to have less dissected fronds with fewer pinnae than terrestrial species (\tab \@ref(tab:pics)), but the range of values for these traits was quite wide at the community level (error bars in \fig \@ref(fig:cwm-div)).
It is likely that the high diversity of non-size related traits we observed reflects different physiological adaptations to various canopy niches.
For example, filmy ferns (family Hymenophyllaceae), which depend on desiccation tolerance to survive dry conditions, have vastly different morphology from bird's nest ferns (*Asplenium nidus* and others), which capture detritus in their rosette leaves.

Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)).
Specific leaf area is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a].
We expected that epiphytes would have lower SLA, given their more extreme growing conditions. 
The lack of a clear difference in SLA between epiphytes and terrestrial species observed here may be an effect of scale: the LES was defined at regional to global scales, but tends to break down at local scales encompassing less environmental variation [@Messier2017].

We did not find that gemmae production in gametophytes was linked with epiphytism. 
Gemmae are asexual propagules that allow gametophytes to persist over multiple growing seasons and attain large population sizes [@Farrar2008].
Gemmae production should be a useful trait in an epiphytic context because it may allow gametophytes to increase their surface area and hedge against population loss due to extreme drying events [@Farrar2008].
Gemmae can also function sexually by producing antherida in the presence of antheridiogen secreted by other gametophytes [@Emigh1977], which may confer a selective advantage in the epiphytic environment by enabling outcrossing [@Dassler2001].
It is possible that some, but not all, epiphytic ferns rely on such strategies, and that they are not under similarly strong selection as other traits such as leaf size and gametophyte morphotype. 
Interestingly, when we classified 'widespread' species as those with gametophytes occurring above or below the minimum or maximum elevational range of conspecific sporophytes using the data of @Nitta2017, we found that gemmae production is correlated with widespread growth (log likelihood dependent model = `r correlated_evo_test_gemmae$dependent.logL %>% round(1) %>% as.numeric`, log likelihood independent model = `r correlated_evo_test_gemmae$independent.logL %>% round(1) %>% as.numeric`, \pval = `r correlated_evo_test_gemmae$P %>% round(3) %>% as.numeric`, Pagel's test of correlated evolution).
Thus, it seems that gemmae may function in long-term gametophyte persistence, but are not required for epiphytic growth.

Presence of hairs and glands in gametophytes has very strong phylogenetic signal (\tab \@ref(tab:phylosig-binary)), and these traits are associated with certain taxonomic groups [@Stokey1951; @Nayar1971].
Although hairs have been hypothesized to reduce rates of drying and increase water uptake after desiccation [@Kappen2007], we did not detect a correlation between hair production and epiphytic growth.
Similarly, production of glands does not appear to be correlated with epiphytic growth.
Wax-secreting glands are prevalent in both sporophytes and gametophytes of many notholaenid ferns, a xeric-adapted clade, but their functional significance has not been established [@Johnson2012].
Further tests are needed to determine if such glands serve a functional role in preventing water loss in desert ferns and epiphytes.

### The role of environmental filtering in the canopy

@@

We obtained partial support for the hypothesis that epiphytic fern communities would be more phylogenetically clustered than terrestrial ones; this was observed for MNTD, but not MPD (\fig \@ref(fig:comm-div)).
This makes sense given that most epiphytes have radiated within five major clades [@Schuettpelz2009]; each epiphytic community we sampled tends to include closely-related species from two or more epiphytic clades, but has a similar overall level of phylogentic diversity as terrestrial communities.
However, we did not detect significant trends in phylogenetic community structure of epiphytic and terrestrial communities along the elevational gradient.
It is possible that the scale of the current study (from \SIrange{200}{1200}{\meter}) prevented the detection of such trends.
When the plots in this study were analyzed together with additional plots from Tahiti reaching a maximum altitude of just over \SI{2000}{\meter}, @Nitta2017 detected increasing phylogenetic clustering in fern sporophyte communities (epiphytes and terrestrial species together) with elevation.
@Kluge2011 also compared phylogenetic and trait diversity between epiphytic and terrestrial ferns over an elevational gradient from \SIrange{100}{3400}{\meter} in Costa Rica, and found trait overdispersion at mid-elevations for epiphytes but no trends in phylogenetic diversity for either group.
They accounted this to evolutionary plasticity in traits, but did not quantify the degree of phylogenetic signal in the traits they studied.

Results for functional community diversity contradicted our hypotheses; epiphytic communities were more functionally diverse than terrestrial as measured by both MPD and MNTD (\fig \@ref(fig:comm-div)).
However, this result is dependent on the traits used to measure functional diversity; when the same analyses were done using only size-related traits (scaled, untransformed stipe length, frond length and width, and rhizome diameter), the opposite result was observed (\fig S1).
Furthermore, terrestrial species show a much wider range of values for size traits (\fig \@ref(fig:cwm-div)), and neither functional group appears to be restricted to a smaller area in trait space (\fig \@ref(fig:pca)).
This suggests that epiphytic ferns are subject to environmental filtering for size-related traits, while simultaneously diversifying along other trait axes.
This could be maintained by niche partitioning within epiphytic habitats, which may be more diverse than terrestrial for a given area [@Rojas-Hernandez2018].
Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk [@Hietz1995].
The vertical stratification of epiphytic species has been well documented in ferns [\eg, @Parra2009; @WatkinsJr.2006a; @Acebey2017] and broader studies [\eg, @Kromer2007; @Zotz2007].
Future research should focus on bringing comparative phylogenetic and community ecological methods to bear on the large body of work characterizing differences within epiphytic microhabitats.

### Conclusion and future directions

We have demonstrated that growth habit, and several traits associated with it, show significant phylogenetic signal in ferns.
Furthermore, we find that epiphytes are constrained by size while being remarkably diverse in other traits.
Finally, we observed drier conditions in the canopy relative to terrestrial habitats. 
Taken together, our results support a scenario of environmental filtering acting to limit size and phylogenetic diversity in epiphytic ferns, while simultaneously enabling diversification along other trait axes.
The generality of this pattern could be further tested by investigating the functional traits of other groups with large epiphytic radiations such as orchids and bromeliads.

Recent studies [e.g., @Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a] have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by @Schuettpelz2009 by failing to detect increased rates of diversification associated with epiphytic growth. 
Epiphytism itself is a complex trait requiring the evolution of multiple associated traits, and the timing and order of the evolution of such traits may be related to the complex diversification dynamics observed in these studies.
Small size seems to be a prerequisite for epiphytic growth, as indicated both by our results and the global study of @Testo2018a.
Here, we have shown that size appears to be the exception, not the rule, when it comes to trait diversity of epiphytes: although epiphytes are universally small, our data suggest they are diverse in most other traits.
This emphasizes that fern epiphytes should not be treated as a monolithic group, but rather encompass a wide range of adaptive forms.
We hope future studies will build on our results by analyzing traits involved in epiphytic growth other than body size across densely sampled phylogenies to dissect how each may relate to diversification and growth habit.
Additionally, future studies should focus on determining in more detail how niche space is partitioned within epiphytic vs. terrestrial habitats, which is clearly a major dividing line for ferns.
